Draft

### 2022-06-19 Final/FinalReport.qmd
edited introduction and research question wording

Understanding public attitudes on divisive political issues is an important way for political leaders to mobilize voters and for lawmakers to draft laws that represent their constituents. While it is simple to poll constituents' positions on an issue, it can be challenging to assess the complex factors that influence and predict those stances.

Abortion is one such divisive issue in the United States, with strong organization of pro-choice and pro-life groups in states across the country in support of legislation for their respective sides (Ziegler, 2020). However, following the Supreme Court's 1973 decision in Roe v. Wade, a ruling that has historically protected women's rights to abortion without excessive government restriction in America, the issue has risen in political salience. Both the pro-choice and pro-life movements have gained national prominence, and the two major political parties have polarized around the issue, with the Democratic Party in favor of and the Republican Party against policies legalization and increased access to abortion (Weinberger 2022). Abortion has also increasingly become a key issue that voters consider, with an increasing share of Americans identifying as "single-issue voters" regarding abortion (Brenan 2020).

In May 2022, a leaked draft opinion revealed that the US Supreme Court is prepared to overturn Roe v. Wade. Overturning Roe would dramatically change the trajectory of abortion politics in the US. Unless the US Congress passed a national policy, states would be able to decide whether or not to legalize abortion and gain much greater leverage in regulating access to the procedure (Weinberger 2022).

Given the potential overturning of Roe and the polarizing nature of the issue, it is important to understand how the American public feels about whether abortion should be legal or not, how accessible the procedure should be, and what factors influence these opinions. Understanding public opinion on the issue will ensure that political leaders are able to mobilize the correct constituencies, and that policy experts are able to pass policies on this issue that accurately reflect the preferences of the American people.

This dataset observes attitudes on the justifiability of abortion among respondents in in the United States across six “waves” of the World Values Survey (1982-2011), which is administered every few years and collects information about people’s values and beliefs worldwide, alongside basic demographic characteristics such as age and college education. The data are collected via face-to-face interviews at the respondents’ homes. In this data set, the response variable, justifiability of abortion, is a numerical measure on a scale of 1 to 10 on the individual person’s attitude toward whether abortion is justifiable or not.  Individuals responded 1 for “abortion is never justified” and 10 for “abortion is always justified.

Our research question is as follows: “What is the relationship between attitudes on the justifiability of abortion and personal attitudes towards other issues, such as political ideology, importance of religion, respect for authority and national pride?” We will attempt to answer this question using an EDA-informed predictive model. Given that this issue has become highly polarized by political party, we predict that liberal political attitudes will correlate with belief that abortion is more justified.

>>>>

Understanding public attitudes on divisive political issues is an important way for political leaders to mobilize voters and for lawmakers to draft laws that represent their constituents. While it may be easier to poll constituents' positions on an issue, it can be challenging to assess the complex factors that influence and predict those stances.

Abortion is one such divisive issue in the United States. Both pro-choice and pro-life groups have a history of mobilizing in states across the country in support of legislation for their respective sides (Ziegler, 2020). However, following the Supreme Court's 1973 decision in Roe v. Wade, a ruling which protected an individual’s right to have an abortion before fetal viability, the issue has risen in political salience. Both the pro-choice and pro-life movements have gained national prominence, and the two major political parties have polarized around the issue, with the Democratic Party in favor of and the Republican Party against policies legalizing and increasing access to abortion (Weinberger 2022). Abortion has also increasingly become a key issue that voters consider when making their choice at the ballot box, with an increasing share of Americans identifying as "single-issue voters" regarding abortion (Brenan 2020).

In May 2022, a leaked draft opinion revealed that the US Supreme Court is prepared to overturn Roe v. Wade. Overturning Roe would dramatically change the trajectory of abortion politics in the US. Unless Congress passes a national policy, states would be able to decide whether or not to legalize abortion and gain much greater leverage in regulating access to the procedure (Weinberger 2022).

Given the potential overturning of Roe and the polarizing nature of the issue, it is important to understand how the American public feels about whether abortion should be legal or not, how accessible the procedure should be, and which factors influence these opinions. Understanding public opinion on the issue will ensure that political leaders are able to mobilize the correct constituencies, and that policy experts are able to pass policies on this issue that accurately reflect the preferences of the American people.

The dataset used in this analysis observes attitudes on the justifiability of abortion among respondents in in the United States across six “waves” of the World Values Survey (1982-2011), which is administered every few years and collects information about people’s values and beliefs worldwide, alongside basic demographic characteristics. The data are collected via face-to-face interviews at the respondents’ homes. In this data set, the response variable, justifiability of abortion, is a numerical measure on a scale of 1 to 10 on the individual person’s attitude toward whether abortion is justifiable or not.  Individuals responded 1 for “abortion is never justified” and 10 for “abortion is always justified.

Our research question is as follows: “Do an individual’s political ideology and their personal attitudes/preferences towards other issues, such as the importance of religion in their life and their respect for authority, among others, predict their attitude on the justifiability of abortion?” We will attempt to answer this question using an EDA-informed predictive model. Given that this issue has become highly polarized by political party, we predict that liberal ideology and liberal-leaning attitudes on other issues will correlate with belief that abortion is more justified.
